---
title: "data-memo"
author: "Haocheng Zhang"
date: "2023-08-20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Read the CSV file into a dataframe
real_estate_data = read.csv('RealEstate_California.csv')

# View the first few rows of the dataframe
head(real_estate_data)
```


## Overview of the Dataset

### What does it include?
- The dataset includes information about real estate listings in California,
such as property details, price, area, number of bedrooms, type of property,
and more.

### Where and how will you be obtaining it?
- [Kaggle: Real Estate California](https://www.kaggle.com/datasets/yellowj4acket/real-estate-california)

### About how many observations? How many predictors?
- Observations (rows): 35,389
- Predictors (columns): 39

### What types of variables will you be working with?
- Numerical variables: price, living area, bathrooms
- Categorical variables: city, state, home type
- Binary variables: hasGarage, isNewConstruction

### Is there any missing data? About how much? Do you have an idea for how to handle it?
- datePostedString: 3 missing values
- time: 289 missing values
- zip-code: 25 missing values (planning to delete these)
- description: 279 missing values
- And i am planning to delete those without zip-code, because it is essential. Also, deleting those 25 data would not impact too much on assurance.

## Overview of Your Research Questions

### What variable(s) are you interested in predicting? What question(s) are you interested in answering?
- I could focus on predicting property prices, depending on different areas and property types.

### Will these questions be best answered with a classification or regression approach?
- Price prediction would typically be a regression problem, while classifying property types would be a classification problem.

### Which predictors do you think will be especially useful?
- Features like living area, bedrooms, bathrooms, location, and property type may be particularly useful.

### Is the goal of your model descriptive, predictive, inferential, or a combination? Explain.
- The primary goal of my model is predictive since I aim to forecast property prices using various features like location (area) and property type.

## Proposed Project Timeline
- Week2: data loading
- Week3: exploratory data analysis
- week4-5: model building, evaluation,
- week6: finalization.

## Any Questions or Concerns
- I am worried that if I should find another dataset for real estate sales in a specific time range, such as 2020-2021, so I can see the trend for price change, which should be helpful to predict future prices. However, the current dataset already includes useful information, such as price, property types, zip-code, which should be already enough for my research.
