---
title: "Untitled"
author: "Haocheng Zhang"
date: "2023-05-28"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Problem 4

#Load the data
polls_data_2016 = read.csv("president_general_polls_sorted_end_date_2016.csv")
polls_data_2020 = read.csv("president_polls_2020.csv")
library(lubridate)
library(usmap)
library(ggplot2)


index_na=which(is.na(polls_data_2020$sample_size)==T)

polls_data_2020=polls_data_2020[which(polls_data_2020$answer==
                                        'Biden'|polls_data_2020$answer==
                                        'Trump'),]

delete_index=which( (polls_data_2020$pollster_id==
                      1610|polls_data_2020$pollster_id==1193)&
                      mdy(polls_data_2020$end_date)!= "2020-11-01")
polls_data_2020=polls_data_2020[-delete_index,]

##now let's delete those poll that only contains one candidate
polls_data_2020_question_id_num=unique(polls_data_2020$question_id)
##clean the data
for(i in 1:length(unique(polls_data_2020$question_id)) ){
  index_set=which(polls_data_2020$question_id==
                    polls_data_2020_question_id_num[i])
  if(length(index_set)!=2){
    polls_data_2020=polls_data_2020[-index_set,]
  }
}

index_NA=which(is.na(polls_data_2020$sample_size)==T)

polls_data_2020=polls_data_2020[-index_NA,]

polls_data_2016_enddate=mdy(polls_data_2016$enddate)

poll_state_sum_clinton_2016=aggregate(polls_data_2016$total.clinton, by=list(State=polls_data_2016$state),FUN=sum)
poll_state_sum_trump_2016=aggregate(polls_data_2016$total.trump, by=list(State=polls_data_2016$state),FUN=sum)

poll_state_diff_percentage=poll_state_sum_clinton_2016
poll_state_diff_percentage[,2]=(poll_state_sum_clinton_2016[,2]-
                                  poll_state_sum_trump_2016[,2])/(poll_state_sum_clinton_2016[,2]+
                                                                    poll_state_sum_trump_2016[,2])
delete_index=which(levels(poll_state_diff_percentage[,1])=='U.S.')
if(length(delete_index)>0){
  poll_state_diff_percentage=poll_state_diff_percentage[-delete_index,]
  poll_state_diff_percentage[,1]
}

state_poll_2016 <- data.frame(
  state =poll_state_diff_percentage[,1],
  diff_percentage=poll_state_diff_percentage[,2]
)

index_biden_2020=which(polls_data_2020$answer=='Biden')
index_trump_2020=which(polls_data_2020$answer=='Trump' )

counts_biden_2020=polls_data_2020$pct[index_biden_2020]*polls_data_2020$
  sample_size[index_biden_2020]
counts_trump_2020=polls_data_2020$pct[index_trump_2020]*polls_data_2020$
  sample_size[index_trump_2020]

polls_data_2020$total.biden=rep(0,dim(polls_data_2020)[1])
polls_data_2020$total.trump=rep(0,dim(polls_data_2020)[1])

polls_data_2020$total.biden[index_biden_2020]=counts_biden_2020
polls_data_2020$total.trump[index_trump_2020]=counts_trump_2020

poll_state_sum_biden_2020=aggregate(polls_data_2020$total.biden, by=list(State=polls_data_2020$state),FUN=sum)
poll_state_sum_trump_2020=aggregate(polls_data_2020$total.trump, by=list(State=polls_data_2020$state),FUN=sum)

poll_state_sum_biden_2020=poll_state_sum_biden_2020[-1,]
poll_state_sum_trump_2020=poll_state_sum_trump_2020[-1,]

state_poll_2020 <- data.frame(
  state =poll_state_sum_biden_2020[,1],
  diff_percentage=(poll_state_sum_biden_2020[,2]-poll_state_sum_trump_2020[,2])/
    (poll_state_sum_biden_2020[,2]+poll_state_sum_trump_2020[,2])
)

limit_val=c(min(state_poll_2016$diff_percentage,state_poll_2020$diff_percentage),
            max(state_poll_2016$diff_percentage,state_poll_2020$diff_percentage))

plot_usmap(data = state_poll_2016, values = "diff_percentage", color = "black") +
  scale_fill_gradient2(name = "difference (%)",   low= "red",
                       mid = "white",
                       high = "blue",
                       midpoint = 0,limits=limit_val)+
  theme(legend.position = "right")+
  ggtitle("2016") 
##2020
##it does not have poll from Nebraska. It does not plot the poll from 
#congressional district
plot_usmap(data = state_poll_2020, values = "diff_percentage", color = "black")+
  scale_fill_gradient2(name = "difference (%)",   low= "red",
                       mid = "white",
                       high = "blue",
                       midpoint = 0,limits=limit_val)+
  theme(legend.position = "right")+
  ggtitle("2020") 
```
```{r}
# Part b
state_poll_2016=state_poll_2016[-c(33,50),]

dif <- data.frame(
  state <- state_poll_2016$state,
  diff_2016 <- state_poll_2016$diff_percentage,
  diff_2020 <- state_poll_2020$diff_percentage
)
index_change = which(dif$diff_2016*dif$diff_2020<0)
dif$state[index_change]
```

```{r}
##part(C)

#No, there are not. The states I've listed that did not flip in reality -- 
#Florida and Nevada -- could have been identified as potential flips due to 
#close polling results, demographic changes, or other state-specific factors. #However, many elements can influence election outcomes beyond what polls 
#predict, such as voter turnout, campaign strategies, and major events.
